---
title: "NowcastPulse: Function Reference and User Guide"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{NowcastPulse: Function Reference and User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # examples not run during build; set eval=TRUE for live runs
)
```

---

# Overview

**NowcastPulse** is a complete nowcasting framework for mixed-frequency
macroeconomic data. It covers every stage of the nowcasting pipeline:

| Stage | Module | Functions |
|---|---|---|
| 1. Data ingestion | Data Loading | `np_load_data()` |
| 2. Seasonal adjustment | Data Processing | `np_seasonal_adjust()` |
| 3. Temporal aggregation | Data Processing | `np_aggregate_monthly()` |
| 4. Dataset merging | Data Processing | `np_merge_monthly()` |
| 5. Full pipeline shortcut | Data Processing | `np_process_data()` |
| 6. Variable transformation | Variable Selection | `np_transform_data()` |
| 7. Correlation ranking | Variable Selection | `np_rank_correlations()` |
| 8. Stepwise variable selection | Variable Selection | `np_select_variables()` |

> **Variable selection, modelling, evaluation, and reporting modules are
> added in subsequent releases.**

---

# Data Format Requirements

Before calling any function, your raw data must be organised as follows.

## Folder Structure

```
<data_dir>/
├── Raw/
│   ├── <prefix>_Daily_Data.csv
│   ├── <prefix>_Daily_Data_Oil.csv
│   ├── <prefix>_Weekly_Data.csv
│   ├── <prefix>_Monthly_Data.csv
│   ├── <prefix>_Quarterly_Data.csv
│   └── <prefix>_Index.csv
└── Processed/          ← created automatically if save_outputs = TRUE
```

## CSV Layout (Data Files)

Each data CSV uses a **wide format** (variables as rows, dates as columns):

| Column position | Content |
|---|---|
| 1st column | `variable_names` — short machine-readable identifier |
| 2nd column | `descriptions` — human-readable label (ignored by package) |
| 3rd column onward | Date labels as column headers; cell values are numeric |

Missing values must be coded as blank cells or `#N/A`.

## Date Header Formats by Frequency

| File | Date format in column headers | Example |
|---|---|---|
| Daily | `YYYYMMDD` | `20000103` |
| Daily Oil | `M/D/YYYY` | `1/3/2000` |
| Weekly | `M/YY/CCYY` (month / 2-digit year) | `1/00/2000` |
| Monthly | `YYYYMM` | `200001` |
| Quarterly | `YYYYQ` (Q = 1–4) | `20001` |

## Index File (`<prefix>_Index.csv`)

Controls seasonal adjustment and temporal aggregation behaviour.
Required columns:

| Column | Values | Meaning |
|---|---|---|
| `variable_names` | string | Must match names in data CSVs |
| `sa_index` | `"NSA"` / anything else | `"NSA"` = apply X-13; otherwise pass through |
| `agg_index` | `"Sum"` / `"Average"` / `"Last"` | How to collapse sub-monthly obs to monthly |

---

# Function Reference

---

## `np_load_data()`

### Purpose

Reads all raw CSV files from `<data_dir>/Raw/` and reshapes each from wide
format (variables × dates) to long format (dates × variables). Also reads
the index file to extract `agg_map` and `sa_map` control vectors.

### Signature

```r
np_load_data(data_dir = "Data/", prefix = "Fiji")
```

### Parameters

| Parameter | Type | Default | Description |
|---|---|---|---|
| `data_dir` | character | `"Data/"` | Root data directory containing a `Raw/` sub-folder |
| `prefix` | character | `"Fiji"` | File-name prefix, e.g. `"Fiji"` → reads `Fiji_Monthly_Data.csv` |

### Return Value

A **named list** with 7 elements:

| Element | Type | Description |
|---|---|---|
| `$daily` | data frame | Daily series; `date` col in `YYYYMMDD` format |
| `$daily_oil` | data frame | Daily oil series; `date` col in `M/D/YYYY` format |
| `$weekly` | data frame | Weekly series; `date` col in `M/YY/CCYY` format |
| `$monthly` | data frame | Monthly series; `date` col in `YYYYMM` format |
| `$quarterly` | data frame | Quarterly series; `date` col in `YYYYQ` format |
| `$agg_map` | named character vector | `variable_name → "Sum" / "Average" / "Last"` |
| `$sa_map` | named character vector | `variable_name → "NSA"` or other flag |

All data frames have `date` as the first column followed by one column per
variable (numeric). The variable `#N/A` string sentinel is converted to `NA`.

### Example

```{r load}
library(NowcastPulse)

raw <- np_load_data(data_dir = "Data/", prefix = "Fiji")

names(raw)
# [1] "daily"     "daily_oil" "weekly"    "monthly"   "quarterly"
#     "agg_map"   "sa_map"

dim(raw$monthly)      # e.g. 304 x 42  (months × variables+date)
raw$agg_map[1:3]      # named vector of aggregation methods
raw$sa_map[1:3]       # named vector of SA flags
```

---

## `np_seasonal_adjust()`

### Purpose

Applies **X-13 ARIMA-SEATS** seasonal adjustment (via the `seasonal`
package) to every variable in all five frequency data frames. Variables
flagged `"NSA"` in `sa_map` are adjusted; all others are passed through
unchanged. Every output column is renamed with an `_SA` suffix.

For non-standard X-13 frequencies (daily `freq = 365`, weekly `freq = 52`),
adjustment is attempted but falls back silently to the original series if
X-13 fails.

### Signature

```r
np_seasonal_adjust(raw_data, sa_map = raw_data$sa_map)
```

### Parameters

| Parameter | Type | Default | Description |
|---|---|---|---|
| `raw_data` | named list | — | Output of `np_load_data()` |
| `sa_map` | named character vector | `raw_data$sa_map` | SA flags per variable; `"NSA"` triggers adjustment |

### Return Value

A **named list** with 5 elements:

| Element | Type | Description |
|---|---|---|
| `$daily` | data frame | `date` + `<var>_SA` columns at daily frequency |
| `$daily_oil` | data frame | `date` + `<var>_SA` columns at daily frequency |
| `$weekly` | data frame | `date` + `<var>_SA` columns at weekly frequency |
| `$monthly` | data frame | `date` + `<var>_SA` columns at monthly frequency |
| `$quarterly` | data frame | `date` + `<var>_SA` columns at quarterly frequency |

Original unadjusted columns are dropped; only `_SA`-suffixed columns remain.

### Example

```{r sa}
sa <- np_seasonal_adjust(raw, sa_map = raw$sa_map)

names(sa)
# [1] "daily" "daily_oil" "weekly" "monthly" "quarterly"

# All columns (except date) end in _SA
grep("_SA$", names(sa$monthly), value = TRUE) |> head(3)
# [1] "gdp_SA"  "cpi_SA"  "trade_SA"
```

> **Note:** X-13 on many variables across daily/weekly series can take
> **5–15 minutes**. Progress is printed variable-by-variable.

---

## `np_aggregate_monthly()`

### Purpose

Converts the daily, daily-oil, and weekly SA data frames to **monthly
frequency** using the per-variable aggregation method from `agg_map`.
The monthly SA data frame is already monthly — its `date` column is
standardised from `"YYYYMM"` to `"YYYY-MM"` and returned as-is.
Quarterly data is excluded from this step (not merged into the monthly file).

### Aggregation methods

| `agg_map` value | Rule |
|---|---|
| `"Sum"` | Sum of all non-NA observations in the month |
| `"Average"` | Mean of all non-NA observations in the month |
| `"Last"` | Last non-NA observation in the month |
| *(not in map)* | Defaults to `"Average"` |

An all-NA month produces `NA` regardless of method.

### Signature

```r
np_aggregate_monthly(sa_data, agg_map = NULL)
```

### Parameters

| Parameter | Type | Default | Description |
|---|---|---|---|
| `sa_data` | named list | — | Output of `np_seasonal_adjust()` |
| `agg_map` | named character vector | `NULL` (→ all Average) | Aggregation method per variable; typically `raw$agg_map` |

### Return Value

A **named list** with 4 elements (quarterly excluded):

| Element | Type | Description |
|---|---|---|
| `$daily` | data frame | Daily SA series aggregated to monthly; `date` = `"YYYY-MM"` |
| `$daily_oil` | data frame | Daily oil SA series aggregated to monthly |
| `$weekly` | data frame | Weekly SA series aggregated to monthly |
| `$monthly` | data frame | Monthly SA series with `date` standardised to `"YYYY-MM"` |

### Example

```{r agg}
monthly_list <- np_aggregate_monthly(sa, agg_map = raw$agg_map)

names(monthly_list)
# [1] "daily"  "daily_oil"  "weekly"  "monthly"

head(monthly_list$daily$date, 3)
# [1] "2000-01" "2000-02" "2000-03"
```

---

## `np_merge_monthly()`

### Purpose

Merges the four monthly data frames from `np_aggregate_monthly()` into a
**single combined data frame** via a full outer join on `date`. Rows before
`start_date` are dropped and the result is sorted chronologically.

This combined data frame is the primary input for all downstream
variable selection and modelling functions.

### Signature

```r
np_merge_monthly(monthly_list, start_date = "2000-01")
```

### Parameters

| Parameter | Type | Default | Description |
|---|---|---|---|
| `monthly_list` | named list | — | Output of `np_aggregate_monthly()` |
| `start_date` | character | `"2000-01"` | Earliest `"YYYY-MM"` date to keep |

### Return Value

A single **data frame**:

| Column | Type | Description |
|---|---|---|
| `date` | character | `"YYYY-MM"` strings, one row per calendar month |
| `<var>_SA` … | numeric | All SA variables from daily, oil, weekly, and monthly sources |

Variables from different frequency sources sit side-by-side. Months with no
observations for a given source appear as `NA` in that variable's column.

### Example

```{r merge}
combined <- np_merge_monthly(monthly_list, start_date = "2000-01")

dim(combined)
# e.g. 304 x 89   (months × total variables across all sources)

combined[1:3, 1:4]
#      date  var1_SA  var2_SA  var3_SA
# 1 2000-01   102.3       NA    0.023
# 2 2000-02   103.1    45.2    0.019
# 3 2000-03   101.8    46.0    0.021
```

---

## `np_process_data()` *(full pipeline wrapper)*

### Purpose

Runs the **complete data processing pipeline** in a single call:

```
np_load_data()  →  np_seasonal_adjust()  →  np_aggregate_monthly()  →  np_merge_monthly()
```

Optionally saves all intermediate and final outputs as CSV files to
`<data_dir>/Processed/`.

### Signature

```r
np_process_data(
  data_dir     = "Data/",
  prefix       = "Fiji",
  start_date   = "2000-01",
  save_outputs = TRUE
)
```

### Parameters

| Parameter | Type | Default | Description |
|---|---|---|---|
| `data_dir` | character | `"Data/"` | Root data directory |
| `prefix` | character | `"Fiji"` | File-name prefix |
| `start_date` | character | `"2000-01"` | Earliest month to retain in the combined output |
| `save_outputs` | logical | `TRUE` | If `TRUE`, writes all processed CSVs to `<data_dir>/Processed/` |

### Output Files Saved (when `save_outputs = TRUE`)

| File | Contents |
|---|---|
| `<prefix>_Daily_Data_SA.csv` | Seasonally adjusted daily data |
| `<prefix>_Daily_Data_Oil_SA.csv` | Seasonally adjusted daily oil data |
| `<prefix>_Weekly_Data_SA.csv` | Seasonally adjusted weekly data |
| `<prefix>_Monthly_Data_SA.csv` | Seasonally adjusted monthly data |
| `<prefix>_Quarterly_Data_SA.csv` | Seasonally adjusted quarterly data |
| `<prefix>_Combined_Monthly_SA.csv` | Final merged monthly dataset |

### Return Value

Returned **invisibly** — a named list with 3 elements:

| Element | Type | Description |
|---|---|---|
| `$sa` | named list (5 data frames) | SA data at native frequency for each source |
| `$monthly` | named list (4 data frames) | SA data aggregated to monthly per source |
| `$combined` | data frame | Final merged monthly dataset (main modelling input) |

### Example

```{r pipeline}
library(NowcastPulse)

result <- np_process_data(
  data_dir     = "Data/",
  prefix       = "Fiji",
  start_date   = "2000-01",
  save_outputs = TRUE
)

# Inspect the combined output
dim(result$combined)
head(result$combined[, 1:6])

# Access intermediate SA results
dim(result$sa$monthly)
dim(result$sa$quarterly)

# Access monthly-aggregated pieces
names(result$monthly)
```

---

# Typical Workflow

```{r workflow}
library(NowcastPulse)

# Option A: one-liner
result <- np_process_data("Data/", prefix = "Fiji")
combined <- result$combined

# Option B: step-by-step (more control)
raw          <- np_load_data("Data/", prefix = "Fiji")
sa           <- np_seasonal_adjust(raw, sa_map = raw$sa_map)
monthly_list <- np_aggregate_monthly(sa, agg_map = raw$agg_map)
combined     <- np_merge_monthly(monthly_list, start_date = "2000-01")

# combined is now ready for variable selection and modelling
str(combined[, 1:5])
```

---

# Internal Helpers (not exported)

These functions are used internally and are not part of the public API.
They can be accessed via `NowcastPulse:::` for debugging.

| Function | Description |
|---|---|
| `wide_to_long(dta)` | Transposes wide CSV format to long (dates × variables) |
| `parse_date_by_type(date_vec, dataset_type)` | Converts raw date strings to `Date` objects by frequency type |
| `adjust_seasonal_align(x, dates, freq)` | Runs X-13 on a single series and re-aligns output to original length |
| `aggregate_to_monthly_internal(df, dataset_type, agg_map)` | Aggregates a data frame to monthly using `tapply` |
| `apply_x13(df, dataset_type, freq, dataset_label, sa_map)` | Applies X-13 to all variables in a data frame |

---

# Session Info

```{r session}
sessionInfo()
```
